---
title: "Installing R"
author: "Mwaura Patrick"
date: "2023-09-14"
output: pdf_document
---

To install R and its essential programs, follow these steps:

## 1. Install R:

- Windows:
   - Visit the CRAN (Comprehensive R Archive Network) website: \textcolor{red}{https://cran.r-project.org/}
   - Choose a nearby CRAN mirror and download the R installer for Windows.
   - Run the installer and follow the installation instructions.

- macOS:
   - You can install R on macOS using the Homebrew package manager or by downloading the installer from the CRAN website.
   - To install using Homebrew, open Terminal and run the following commands:
   
```{r}
##/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/master/install.sh)"
##brew install --cask r

```

- Linux (Ubuntu as an example):
  - Open a terminal and run the following commands:

```{r}
##sudo apt update
##sudo apt install r-base

```


## 2. Install RStudio (optional but highly recommended):

- RStudio is a popular integrated development environment (IDE) for R that makes working with R much easier. You can download it from the RStudio website: \textcolor{red} {https://www.rstudio.com/products/rstudio/download/}

## 3. Install essential R packages:
   - Once you have R installed, open R or RStudio. You can install R packages (libraries) using the `install.packages()` function. For example, to install the popular `dplyr` package, open R or RStudio and run:
   
```{r}
#install.packages("dplyr")

```

- Install any other packages you need for your specific tasks by replacing `"dplyr"` with the package name.

## 4. Start using R:
- You can now start using R by opening R or RStudio. You can write and run R scripts, interactively work in the console, and perform data analysis and visualization.

## 5. Learn R:
- To effectively use R, consider learning the language and its data analysis capabilities. You can find many tutorials, courses, and books online to help you get started.

