---
title: "Introduction to R"
author: "Mwaura Patrick"
date: "2023-09-13"
output: pdf_document
---

# Introduction

This document provides an introduction to the R programming language. It covers various features of the R environment and demonstrates how to use them.

## Sample Session

In this section, we will explore some basic features of the R environment through sample sessions. These examples will help you become familiar with R's capabilities.

```{r}
# Help Start
# help.start()
```

Start the HTML interface to online help using your web browser. Explore the features of this facility to get acquainted with it.

## Generate Random Vectors

Let's generate two pseudo-random normal vectors for x- and y-coordinates.

```{r}
x <- rnorm(50)
y <- rnorm(x)
```

## Plot Data

Plot the points in the plane, and a graphics window will appear automatically.

```{r}
plot(x, y)
```

## R Workspace

Check which R objects are currently in the R workspace.

```{r}
ls()
```

## Data Cleanup

Remove objects that are no longer needed to clean up the workspace.

```{r}
rm(x, y)
```

## Creating Data

Make x a sequence from 1 to 20.

```{r}
x <- 1:20
```

## Weight Vector

Create a 'weight' vector of standard deviations.

```{r}
w <- 1 + sqrt(x) / 2
```

## Data Frame

Create a data frame of two columns, x and y, and display it.

```{r}
dummy <- data.frame(x = x, y = x + rnorm(x) * w)
dummy
```

## Linear Regression

Fit a simple linear regression model and view the analysis.

```{r}
fm <- lm(y ~ x, data = dummy)
summary(fm)
```

## Weighted Regression

Perform a weighted regression.

```{r}
fm1 <- lm(y ~ x, data = dummy, weight = 1 / w^2)
summary(fm1)
```

## Variables from Data Frame

Make the columns in the data frame visible as variables and can be accessed by simply giving their names.

Usage

```{r}
attach(dummy)
```

## Nonparametric Local Regression

Create a nonparametric local regression function.

```{r}
lrf <- lowess(x, y)
```

## Plotting

Generate standard point plots.

```{r}
plot(x, y)
```

## Adding Local Regression

Add the local regression to the plot.

```{r}
plot(x, y)
lines(x, lrf$y)
```

## True Regression Line

Include the true regression line.

```{r}
plot(x, y)
lines(x, lrf$y)
abline(0, 1, lty = 3)
```

## Unweighted Regression Line

Plot the unweighted regression line.

```{r}
plot(x, y)
lines(x, lrf$y)
abline(0, 1, lty = 3)
abline(coef(fm))
```

## Weighted Regression Line

Plot the weighted regression line.

```{r}
plot(x, y)
lines(x, lrf$y)
abline(0, 1, lty = 3)
abline(coef(fm))
abline(coef(fm1), col = "red")
```

## Data Frame Cleanup

Remove the data frame from the search path.

```{r}
detach()
```

## Regression Diagnostic Plot

Create a standard regression diagnostic plot to check for heteroscedasticity.

```{r}
plot(fitted(fm), resid(fm),
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
```

## Normal Scores Plot

Generate a normal scores plot to check for skewness, kurtosis, and outliers.

```{r}
qqnorm(resid(fm), main = "Residuals Rankit Plot")
```

## Cleanup

Clean up the workspace.

```{r}
rm(fm, fm1, lrf, x, dummy)
```

## Michelson Classical Experiment

The next section explores data from the classical experiment of Michelson to measure the speed of light.

## Data File Path

Get the path to the data file.

```{r}
filepath <- system.file("data", "morley.tab", package = "datasets")
filepath
```

## Optional File Review

Optionally, you can review the data file.

```{r}
# file.show(filepath)
```

## Read Data

Read in the Michelson data as a data frame and examine it.

```{r}
mm <- read.table(filepath)
head(mm)
tail(mm)
```

## Change Data Types

Change Expt and Run into factors.

```{r}
mm$Expt <- factor(mm$Expt)
mm$Run <- factor(mm$Run)
```

## Attach Data Frame

Make the data frame visible at position 2 (the default).

```{r}
attach(mm)
```

## Boxplots

Compare the five experiments with simple boxplots.

```{r}
plot(Expt, Speed, main = "Speed of Light Data", xlab = "Experiment No.")
```

## Analyze Data

Analyze the data as a randomized block, with 'runs' and 'experiments' as factors.

```{r}
fm <- aov(Speed ~ Run + Expt, data = mm)
summary(fm)
```

## Sub-Model Analysis

Fit a sub-model omitting 'runs' and compare using a formal analysis of variance.

```{r}
fm0 <- update(fm, . ~ . - Run)
anova(fm0, fm)
```

## Clean Up

Clean up before moving on.

```{r}
detach()
rm(fm, fm0)
```

# Graphical Features

## Contour and Image Plots

In this section, we explore contour and image plots.

## Data Preparation

Prepare data for plotting.

```{r}
x <- seq(-pi, pi, len = 50)
y <- x
```

## Create a Matrix

Create a square matrix of values for the function cos(y) / (1 + x^2).

```{r}
f <- outer(x, y, function(x, y) cos(y) / (1 + x^2))
```

## Set Plotting Parameters

Set the plotting parameters and the plotting region to "square."

```{r}
oldpar <- par(no.readonly = TRUE)
par(pty = "s")
```

## Contour Map

Make a contour map of f and add more lines for detail.

```{r}
contour(x, y, f)
contour(x, y, f, nlevels = 15, add = TRUE)
```

## Asymmetric Part

Calculate the "asymmetric part" of f.

```{r}
fa <- (f - t(f)) / 2
```

## Contour Plot

Make a contour plot.

```{r}
contour(x, y, fa, nlevels = 15)
```

## Restore Graphics Parameters

Restore the old graphics parameters.

```{r}
par(oldpar)
```

## Image Plots

Create high-density image plots.

```{r}
image(x, y,

 f)
image(x, y, fa)
```

## Clean Up

Clean up before moving on.

```{r}
objects()
rm(x, y, f, fa)
```

# Complex Arithmetic

In this section, we explore complex arithmetic in R.

## Complex Numbers

Create complex numbers.

```{r}
th <- seq(-pi, pi, len = 100)
z <- exp(1i * th)
```

## Complex Arguments Plot

Plot complex arguments.

```{r}
par(pty = "s")
plot(z, type = "l")
```

## Sampling Points

Sample points within the unit circle using two methods.

```{r}
# Method 1
w <- rnorm(100) + rnorm(100) * 1i
w <- ifelse(Mod(w) > 1, 1 / w, w)

# Method 2
w <- sqrt(runif(100)) * exp(2 * pi * runif(100) * 1i)
```

## Plot Sampling Results

Plot the sampled points.

```{r}
plot(w, xlim = c(-1, 1), ylim = c(-1, 1), pch = "+", xlab = "x", ylab = "y")
lines(z)
```

## Cleanup

Clean up the workspace.

```{r}
rm(th, w, z)
```

## Quit R

You will be asked if you want to save the R workspace. For an exploratory session like this, you probably do not want to save it.

```{r}
# q()
```

This document provides a comprehensive introduction to R, covering various aspects of the language and its capabilities.
```

This revised document is better organized, making it easier to read and follow. Each section is clearly labeled, and code blocks are appropriately formatted. Additionally, I added some headings to provide structure and improve readability.